A comparison of design-based and model-based approaches for finite population spatial data.

### 2022-04-11 inst/manuscript/manuscript.tex
minor updates

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualations and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between
    these approaches is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used. This suggests that
    the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of population units according to a
sampling design. A sampling design assigns a positive probability of
inclusion (inclusion probability) in the sample to each population unit.
These inclusion probabilities are later used to estimate population
parameters. Some examples of commonly used sampling designs include
simple random sampling, stratified random sampling, and cluster
sampling.

When sampling designs incorporate spatial locations into sampling, we
call the resulting samples ``spatially balanced.'' One approach to
selecting spatially balanced samples is the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004),
which we discuss in more detail in Section \ref{subsec:spb_design}. When
sampling designs do not incorporate spatial locations into sampling, we
call the resulting samples ``non-spatially balanced.''

@@

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of a
fixed, unknown mean, we are the value of the mean, a random variable.
Prediction intervals are then derived using assumptions of the
data-generating stochastic process. If we repeatedly generate response
values from the same process and select samples, then 95\% of all 95\%
prediction intervals constructed from a procedure with appropriate
coverage will contain their respective realized means. Cressie (1993)
and Schabenberger and Gotway (2017) provide thorough reviews of
model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we
provide a visual comparison of the design-based and model-based
approaches (Ver Hoef (2002) and Brus (2021) provide similar figures).

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples (samples that incorporate spatial
locations of the population units). Spatially balanced samples are
useful because parameter estimates from these samples tend to vary less
than parameter estimates from samples that are not spatially balanced
(Barabesi and Franceschi, 2011; Benedetti et al., 2017; Grafström and
Lundström, 2013; Robertson et al., 2013; Stevens and Olsen, 2004; Wang
et al., 2013). The first spatially balanced sampling algorithm to see
widespread use was the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004). To quantify the spatial balance of
a sample, Stevens and Olsen (2004) proposed loss metrics based on
Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was
developed, several other spatially balanced sampling algorithms emerged,
including the Local Pivotal Method (Grafström et al., 2012; Grafström
and Matei, 2018), Spatially Correlated Poisson Sampling (Grafström,
2012), Balanced Acceptance Sampling (Robertson et al., 2013),
Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning Sampling (Robertson et al., 2018). In this
manuscript, we select spatially balanced samples using the Generalized
Random Tessellation Stratified (GRTS) algorithm because it is readily
available in the \texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle
et al., 2022) and naturally accommodates finite and infinite sampling
frames, unequal inclusion probabilities, and replacement units
(replacement units are population units that can be sampled when a
population unit originally selected can no longer be sampled).

@@

After determining \(\mathcal{A}_k\), it is placed into hierarchical
order. Hierarchical order is a numeric order that first sorts
\(\mathcal{A}_k\) by the level-one addresses from smallest to largest,
then sorts \(\mathcal{A}_k\) by the level-two addresses from smallest to
largest, and so on. For example, \(\mathcal{A}_2\) in hierarchical order
is the set
\(\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}\). Because
hierarchical ordering sorts by level-one cells, then level-two cells,
and so on, population units that have similar hierarchical addresses
tend to be nearby one another in space.Next each population unit is
mapped to a one-dimensional line in hierarchical order where each
population unit's inclusion probability equals its line-length. If a
level-\(k\) cell has multiple population units in it, they are randomly
placed within the cell's respective line segment. A uniform random
variable is then simulated in \([0, 1]\) and a systematic sample is
selected on the line, yielding \(n\) sample points for a sample size
\(n\). Each element in this systematic sample falls on some population
unit's line segment, and thus that population unit is selected in the
sample.

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. Via this mapping, units in two-dimensional
space are partitioned using a hierarchical address. This hierarchical
address is used to map population units to a one-dimensional line. On
the one dimensional line, each population unit's line length equals its
inclusion probability. Then, a systematic sample of population units is
selected on the line and mapped back to two-dimensional space, yielding
the desired sample. Stevens and Olsen (2004) provide more technical
details.

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically.
Second, these estimators ignore the spatial locations of the population
units. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) and yield more precise (narrower) confidence intervals
compared to variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next -- for technical details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the distance between population units), isotropic (independent of
direction), and decay with distance between population units (Cressie,
1993). Henceforth, it is implied that we have made these same
assumptions regarding \(\bm{\delta}\), though Chiles and Delfiner
(1999), pp.~80-93 discuss covariance functions that are not second-order
stationary, not isotropic, or not either. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (Cressie
(1993) provides a thorough list of spatial covariance functions). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in Ver Hoef (2008), we
note here that the predictor and its variance are both moment-based,
meaning that they do not rely on any distributional assumptions.
Distributional assumptions are used, however, when constructing
prediction intervals.

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013) and random forest (Breiman, 2001), among
others, could also be used to obtain predictions for a mean or total
from finite population spatial data. Compared to the k-nearest-neighbors
and random forest approach, we prefer FBPK because it is model-based and
relies on theoretically-based variance estimators leveraging the model's
spatial covariance structure, whereas k-nearest-neighbors and random
forests use ad-hoc variance estimators (Ver Hoef and Temesgen, 2013).
Additionally, Ver Hoef and Temesgen (2013) studied compared FBPK,
k-nearest-neighbors, and random forest in a variety of spatial data
contexts, and FBPK tended to perform best.

@@

\hypertarget{sec:mm_sim}{%
\subsection{Simulation Study}\label{sec:mm_sim}}

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
was IRS-Design. In IRS-Design, samples were selected with the
Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores
the spatial locations of the population units, thus the IRS samples were
not spatially balanced. In IRS-Design, samples were analyzed using the
design-based approach via the Horvitz-Thompson mean estimator and an IRS
variance estimator that ignored the spatial locations of the units in
the sample. The second sampling-analysis combination was IRS-Model,
where samples were selected with the IRS algorithm and analyzed using
the model-based approach via Restricted Maximum Likelihood (REML)
estimation (Harville, 1977; Patterson and Thompson, 1971; Wolfinger et
al., 1994). The third sampling-analysis combination was GRTS-Design,
where samples were selected with the GRTS algorithm and analyzed using
the design-based approach via the Horvitz-Thompson mean estimator and
the local neighborhood variance estimator (which does incorporate the
spatial locations of the units in the sample). The fourth and final
sampling-analysis combination was GRTS-Model, where samples were
selected with the GRTS algorithm and analyzed using the model-based
approach via REML estimation. These sampling-analysis combinations are
also provided in Table \ref{tab:designanalysis}. Lastly we note that for
both the IRS and GRTS samples, equal inclusion probabilities were
assumed for all population units. When IRS assumes equal inclusion
probabilities for all population units, the algorithm is equivalent to
simple random sampling (SRS).

@@

\hypertarget{sec:mm_app}{%
\subsection{Application}\label{sec:mm_app}}

@@

\hypertarget{sec:r_sim}{%
\subsection{Simulation Study}\label{sec:r_sim}}

@@

\hypertarget{sec:r_app}{%
\subsection{Application}\label{sec:r_app}}

@@

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
also more naturally handles binary data, free from the more complicated
logistic regression framework commonly used to analyze binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable. The model-based approach also
yields estimated spatial covariance parameters, which help better
understand the dependence structure in the stochastic process of study.
Model selection is also possible using model-based approaches and
criteria such as cross validation, likelihood ratio tests, or AIC
(Akaike, 1974). Model-based approaches are capable of more efficient
small-area estimation than design-based approaches by leveraging
distributional assumptions in areas with few observed units. Model-based
approaches can also compute unit-by-unit predictions at unobserved
locations and use them to construct informative visualizations like
smoothed maps. In short, when deciding whether the design-based or
model-based approach is more appropriate to implement, the benefits and
drawbacks of each approach should be considered alongside the particular
goals of the study.

@@

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data. In the simulated data, a variety of sample sizes, location
    layouts, dependence structures, and response types are considered. In
    the simulated and real data, the population mean is the parameter of
    interest and performance is measured using statistics like bias,
    squared error, and interval coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the Generalized Random
    Tessellation Stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    Simple Random Sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based approaches tend to outperform design-based approaches,
    even for skewed data where the model-based distributional assumptions
    are violated. The performance gap between these approaches is small
    GRTS samples are used but large when SRS samples are used. This
    suggests that the sampling choice (whether to use GRTS or SRS) is most
    important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of population units according to a
sampling design. A sampling design assigns a non-zero probability of
inclusion (inclusion probability) in the sample to each population unit.
These inclusion probabilities are later used to estimate population
parameters. Some examples of commonly used sampling designs include
simple random sampling, stratified random sampling, and cluster
sampling.

When samples are chosen in a manner such that the layout of sampled
units reflects the layout of the population units, we call the resulting
sample ``spatially balanced.'' By ``reflecting the layout of the
population units'', we mean that if population units are concentrated in
specific areas, the units in the sample should be concentrated in the
same areas. Because spatially balanced samples reflect the layout of the
population units, they are not necessarily ``spread out'' in space in
some equidistant manner.

One approach to selecting spatially balanced samples is the Generalized
Random Tessellation Stratified (GRTS) algorithm (Stevens and Olsen,
2004), which we discuss in more detail in Section
\ref{subsec:spb_design}. When sampling designs do not incorporate
spatial locations into sampling, we call the resulting samples
``non-spatially balanced.''

@@

Instead of estimating fixed, unknown population parameters, as in the
design-based approach, often the goal of model-based inference is to
predict a realized variable, or value. For example, suppose the realized
mean of all population units is the value of interest. Instead of a
fixed, unknown mean, we are predicting the value of the mean, a random
variable. Prediction intervals are then derived using assumptions of the
data-generating stochastic process. If we repeatedly generate response
values from the same process and select samples, then 95\% of all 95\%
prediction intervals constructed from a procedure with appropriate
coverage will contain their respective realized means. Cressie (1993)
and Schabenberger and Gotway (2017) provide thorough reviews of
model-based approaches for spatial data. In Fig. \ref{fig:fig1}, we
provide a visual comparison of the design-based and model-based
approaches (Ver Hoef (2002) and Brus (2021) provide similar figures).

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less than
parameter estimates from samples that are not spatially balanced
(Barabesi and Franceschi, 2011; Benedetti et al., 2017; Grafström and
Lundström, 2013; Robertson et al., 2013; Stevens and Olsen, 2004; Wang
et al., 2013). The first spatially balanced sampling algorithm to see
widespread use was the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004). To quantify the spatial balance of
a sample, Stevens and Olsen (2004) proposed loss metrics based on
Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was
developed, several other spatially balanced sampling algorithms emerged,
including the Local Pivotal Method (Grafström et al., 2012; Grafström
and Matei, 2018), Spatially Correlated Poisson Sampling (Grafström,
2012), Balanced Acceptance Sampling (Robertson et al., 2013),
Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning Sampling (Robertson et al., 2018). In this
manuscript, we select spatially balanced samples using the Generalized
Random Tessellation Stratified (GRTS) algorithm because it is readily
available in the \texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle
et al., 2022) and naturally accommodates finite and infinite sampling
frames, unequal inclusion probabilities, and replacement units
(replacement units are population units that can be sampled when a
population unit originally selected can no longer be sampled).

@@

After determining \(\mathcal{A}_k\), it is placed into hierarchical
order. Hierarchical order is a numeric order that first sorts
\(\mathcal{A}_k\) by the level-one addresses from smallest to largest,
then sorts \(\mathcal{A}_k\) by the level-two addresses from smallest to
largest, and so on. For example, \(\mathcal{A}_2\) in hierarchical order
is the set
\(\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}\). Because
hierarchical ordering sorts by level-one cells, then level-two cells,
and so on, population units that have similar hierarchical addresses
tend to be nearby one another in space.Next each population unit is
mapped to a one-dimensional line in hierarchical order where each
population unit's inclusion probability equals its line-length. If a
level-\(k\) cell has multiple population units in it, they are randomly
placed within the cell's respective line segment. A uniform random
variable is then simulated in \([0, 1]\) and a systematic sample is
selected on the line, yielding \(n\) sample points for a sample size
\(n\). Each element in this systematic sample falls on some population
unit's line segment, and thus that population unit is selected in the
sample. For further details regarding the GRTS algorithm, see Stevens
and Olsen (2004).

@@

It is also important to quantify the uncertainty in \(\hat{\tau}_{ht}\).
Horvitz and Thompson (1952) and Sen (1953) provide variance estimators
for \(\hat{\tau}_{ht}\), but these estimators have two drawbacks. First,
they rely on calculating \(\pi_{ij}\), the probability that population
unit \(i\) and population unit \(j\) are both in the sample -- this
quantity can be challenging if not impossible to calculate analytically
for GRTS samples. Second, these estimators ignore the spatial locations
of the population units. To address these two drawbacks simultaneously,
Stevens and Olsen (2003) proposed the local neighborhood variance
estimator. The local neighborhood variance estimator does not rely on
\(\pi_{ij}\) and estimates the variance of \(\hat{\tau}\) conditional on
the random properties of the GRTS sample -- the idea being that this
conditioning should yield a more precise estimate of \(\hat{\tau}\).
They show that the each observation's contribution to the overall
variance is dominated by local variation. Thus the local neighborhood
variance estimator is a weighted sum of variance estimates from each
observation's local neighborhood. These local neighborhoods contain
observation itself and its three nearest neighbors. For more details,
see Stevens and Olsen (2003).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- for technical details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the distance between population units), isotropic (independent of
direction), and decay with distance between population units (Cressie,
1993). Henceforth, it is implied that we have made these same
assumptions regarding \(\bm{\delta}\), though Chiles and Delfiner
(1999), pp.~80-93 discuss covariance functions that are not second-order
stationary, not isotropic, or not either. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (Cressie
(1993) provides a thorough list of spatial covariance functions). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent variability, \(\sigma^2_{2}\) is the
variance parameter the quantifies that spatially independent
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). In geostatistical
literature, \(\sigma^2_{1}\) is often called the partial sill and
\(\sigma^2_{2}\) is often called the nugget.

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on using restricted maximum
likelihood (Harville, 1977; Patterson and Thompson, 1971; Wolfinger et
al., 1994). REML is preferred over maximum likelihood (ML) because ML
estimates can be badly biased for small sample sizes, due to the fact
that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\delta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\delta}}_{reml}\), the REML estimates of \(\bm{\delta}\).
Then \(\bm{\beta}_{reml}\), the REML estimate of \(\bm{\beta}\), is
given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\delta}}_{reml}\).

With the model formulation in Equation\(~\)\ref{equation:Zmarginal}, the
Best Linear Unbiased Predictor (BLUP) for \(f(\mathbf{b}'\mathbf{z})\)
and its prediction variance can be computed. While details of the
derivation are in Ver Hoef (2008), we note here that the predictor and
its variance are both moment-based, meaning that they do not rely on any
distributional assumptions. Distributional assumptions are used,
however, when constructing prediction intervals.

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013) and random forest (Breiman, 2001), among
others, could also be used to obtain predictions for a mean or total
from finite population spatial data. Compared to the k-nearest-neighbors
and random forest approach, we prefer FPBK because it is model-based and
relies on theoretically-based variance estimators leveraging the model's
spatial covariance structure, whereas k-nearest-neighbors and random
forests use ad-hoc variance estimators (Ver Hoef and Temesgen, 2013).
Additionally, Ver Hoef and Temesgen (2013) compared FPBK,
k-nearest-neighbors, and random forest in a variety of spatial data
contexts, and FPBK tended to perform best.

@@

\hypertarget{sec:mm_sim}{%
\subsection{Simulated Data}\label{sec:mm_sim}}

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
was IRS-Design. In IRS-Design, samples were selected with the
Independent Random Sampling (IRS) algorithm. The IRS algorithm ignores
the spatial locations of the population units, thus the IRS samples were
not spatially balanced. In IRS-Design, samples were analyzed using the
design-based approach via the Horvitz-Thompson mean estimator and an IRS
variance estimator that ignored the spatial locations of the units in
the sample. The second sampling-analysis combination was IRS-Model,
where samples were selected with the IRS algorithm and analyzed using
the model-based approach while estimating the covariance parameters
(\(\bm{\delta}\)) and fixed effects (\(\bm{\beta}\) using restricted
maximum likelihood (REML). The third sampling-analysis combination was
GRTS-Design, where samples were selected with the GRTS algorithm and
analyzed using the design-based approach via the Horvitz-Thompson mean
estimator and the local neighborhood variance estimator (which does
incorporate the spatial locations of the units in the sample). The
fourth and final sampling-analysis combination was GRTS-Model, where
samples were selected with the GRTS algorithm and analyzed using the
model-based approach while estimating the covariance parameters
(\(\bm{\delta}\)) and fixed effects (\(\bm{\beta}\) using restricted
maximum likelihood (REML). These sampling-analysis combinations are also
provided in Table \ref{tab:designanalysis}. Lastly we note that for both
the IRS and GRTS samples, equal inclusion probabilities were assumed for
all population units. When IRS assumes equal inclusion probabilities for
all population units, the algorithm is equivalent to simple random
sampling (SRS).

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/symm_pop_hist.jpeg}
  \caption{}
  \label{fig:symm_pop_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/skew_pop_hist.jpeg}
  \caption{}
  \label{fig:skew_pop_hist}
\end{subfigure}
\caption{Histograms of an example population realization for the normal (a) and skewed (b) scenarios for the simulated data.}
\label{fig:sim_pops}
\end{figure}

\hypertarget{sec:mm_app}{%
\subsection{National Lakes Assessment Data}\label{sec:mm_app}}

@@

\hypertarget{sec:r_sim}{%
\subsection{Simulated Data}\label{sec:r_sim}}

@@

\hypertarget{sec:r_app}{%
\subsection{National Lakes Assessment DAta}\label{sec:r_app}}

@@

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
easily handles binary data through a straightforward application of the
Horvitz-Thompson estimator. In contrast, analyzing binary data using a
model-based approach generally requires a logistic mixed regression
model, which can be challenging to estimate and interpret (Bolker et
al., 2009). The model-based approach, however, can more naturally
quantify the relationship between covariates (predictor variables) and
the response variable. The model-based approach also yields estimated
spatial covariance parameters, which help better understand the
dependence structure in the stochastic process of study. Model selection
is also possible using model-based approaches and criteria such as cross
validation, likelihood ratio tests, or AIC (Akaike, 1974). Model-based
approaches are capable of more efficient small-area estimation than
design-based approaches by leveraging distributional assumptions in
areas with few observed units. Model-based approaches can also compute
unit-by-unit predictions at unobserved locations and use them to
construct informative visualizations like smoothed maps. In short, when
deciding whether the design-based or model-based approach is more
appropriate to implement, the benefits and drawbacks of each approach
should be considered alongside the particular goals of the study.

@@

\leavevmode\hypertarget{ref-bolker2009generalized}{}%
Bolker, B.M., Brooks, M.E., Clark, C.J., Geange, S.W., Poulsen, J.R.,
Stevens, M.H.H., White, J.-S.S., 2009. Generalized linear mixed models:
A practical guide for ecology and evolution. Trends in ecology \&
evolution 24, 127--135.
